---
title: "Day 3 Practice"
output: html_document
---

0. Make a "copy" of the `iris` data set in your workspace.

```{r}
myiris <- iris
```

1. Create a new data frame that is the subset of rows where Sepal Width is equal to or greater than 3.3.  

```{r}
myiris_sep = subset(myiris, myiris$Sepal.Width >= 3.3)

#scott's way: myris.sub1 = myris[myris$Petal.Length >= 3.3, ]
#to combine: myris.sub1 = myris [myris$Petal.Length >= 3.3 & myris$Sepal.Width > 2.5, ]

```

2. Create a new data frame that is the subset of rows equal to or less than the median value of Petal Width.

```{r}
med_width = median(myiris$Petal.Width)
myiris_wid = subset(myiris, myiris$Petal.Width <= med_width)

```

3. Use `sample()` to "scramble" the data set. HINT: think about how to create a vector of numbers that represents a new ordering of the current rows.

```{r}

scrambled = myiris[sample(nrow(myiris)), ]


```

4. Using `summary()`, identify a factor in the iris data

```{r}
summary(myiris)
#species is a factor
```

5. Use subsetting to create a data frame of just the virginica species, then re-run `summary()` to check. Then "clean up" the Species factor so that it only has one level. Show how this works for both the original and the "scrambled" data frame.

```{r}
virginica = subset(myiris, myiris$Species == "virginica")
summary(virginica)

virginica$Species = factor(virginica$Species)
summary(virginica$Species)
#------

virginica_scram = subset(scrambled, scrambled$Species == "virginica")
summary(virginica_scram)

virginica_scram$Species = factor(virginica_scram$Species)
summary(virginica_scram$Species)

```

6. In the full data set (your copy), create different versions of the Species column:
   a. a numeric vector
   b. a character vector
   
```{r}
myiris$species.num = as.numeric(myiris$Species)
myiris$species.char = as.character(myiris$Species)

```

7. Use `xtabs()` to see how the values in these new columns line up with the original factor column.

```{r}
####NOT SURE ABOUT THIS ONE
xtabs(myiris$Species ~ myiris$species.num)
      

```

8. Create another column that is a copy of the factor column. Now, recode this new column so that the factor levels have new names. 

```{r}
myiris$Species2 = myiris$Species


```

9. Check the order of levels of Species with `levels()`. Create a new column from this factor, changing the order of the levels, using:
  a. `relevel()`
  b. `reorder()`
  c. `factor()`

```{r}

```

10. Explore coersion! Try to make a series of vectors with combinations from each of these data types:
  - numeric
  - character
  - factor
  - logical

For each combination, see what happens, by using `typeof()`, `class()` and `print()` before and after.

```{r}

```

11. Make a new data frame from scratch, with the following structure:
  - a numeric "ID" column that goes from 1 to 20, repeating twice (HINT: see `?seq`)
  - a factor "treatment" column that is "a" or "b", such that each ID matches up with both "a" and "b" once each
  - a numeric "outcome" variable that is a normally distributed variable, such that the outcomes for "b" are 1 unit higher on average 

```{r}

```

12. Now turn your ID column into a factor in the following way:
  - append "s" at the beginning of each number
  - convert into a factor
  - check the order of levels. Do you care? How might you change the order?

```{r}

```

13. Now change the "treatment" factor to replace "a" with "old practice" and "b" with "new technique". Do this carefully by first creating a new column that is a straight copy of the "treatment" column, then modifying that new column, then (using `xtabs()`) confirm that your new coding matches up with the old one.

```{r}

```

14. Now add a new column that just picks out "extreme values", aka values of "outcome" that are > the 90th percentile

```{r}

```

