---
title: "Day 8 Lecture"
output: html_document
---

# Today's agenda
  - questions/review
  - if/else
  - for-loops
    - counters (sets)
    - "growing" results
    - debugging
    - break, next
  - simulating data
  - "roll your own" power analysis

# Questions/review?

```{r}
#question about factors, and releveling, etc

mydata = data.frame(n = 1:10, 
                   measure = rnorm(10), 
                   myfactor = c(rep("pre", 4), rep("post", 6)), 
                   stringsAsFactors = FALSE)

head(mydata)
summary(mydata)

mydata$myfactor2 = as.factor(mydata$myfactor)
summary(mydata)
levels(mydata$myfactor2)

mydata$myfactor3 = factor(mydata$myfactor, levels = c("pre", "post"))

mydata$myfactor4 = factor(mydata$myfactor, levels = c("pre", "post", "something else"))
summary(mydata)

```



# if/else
  - form: if(LOGICAL TEST) { CODE BLOCK if TEST == TRUE  }
  -       else if(OPTIONAL ADDITIONAL TEST) { CODE BLOCK if TRUE }
  - "else" code only tested if initial "if" fails
  - NOTE: "else" MUST be on same line as "}" that closes initial "if" block

```{r}
# simple examples
# first curly bracket has to be on the same line as the if. usualy you indent the first line after the if
# this is all style but R does some of it by default
# careful that what you put in the if() only results in ONE T/F, not a whole vector

x = 3

if(x < 5) {
  print("x is small")
}


x = 6

if(x < 5) {
  print("x is small")
}


#Here, white space matters!!!!  'else' has to be on the same line as the closing } for the if statement

x = 6

if(x < 5) {
  print("x is small")
} else {
  print("x must be large")
}


# Can keep adding elses and ifs


# Else really only applies to the previous if
# next if has to be OUTSIDE of braces
# things in braces are the code you want to HAPPEN, outside braces is conditional

x = 5

if(x < 5) {
  print("x is small")
} else if (x > 5) {
  print("x is too large")
} else {
  print ("x is just right")
}


#what about NAs??

x = NA

if(is.na(x)) {
  print("x is missing, go find an x!")
} else if(x < 5) {
  print("x is small")
} else if (x > 5) {
  print("x is too large")
} else {
  print ("x is just right")
}

# have to put the NA condition first, otherwise you're in trouble.
# once an if returns true, all following elses are totally ignored

# what if you had a vector as x? it only checks the first value, and throws a warning.

x = 1:6

if(is.na(x)) {
  print("x is missing, go find an x!")
} else if(x < 5) {
  print("x is small")
} else if (x > 5) {
  print("x is too large")
} else {
  print ("x is just right")
}

```

  - compare with `ifelse()`
    - arguments are: test = [LOGICAL TEST], yes = [VALUE if TEST == TRUE], no = [VALUE if TEST == FALSE]

```{r} 
#ifelse as a function, allows you to work with a vector - i.e. comparing vectors to another, or something

x = 1:10
ifelse(x %% 2 == 0, "even", "odd")   ## %% is "modulo" means remainder of a division. so if the remainder is 0 it's even

ifelse(x %% 2 == 0, "even",
       ifelse (x %% 3 == 0, "divisible by three", "some other odd")) #can nest ifelse's

#ifelse() is more useful for bigger things. 

```

# PRACTICE #1 & #2

# `for` loops
  - simple way to iterate through a set of code, via a "counter" variable
  - avoid re-writing code, when possible!

```{r}
# loops let you re-use code! woo hoo!

# simple vectorized - the plus 2 applies to everything in the vector.
x = 1:10
x + 2


# basic structure of a loop
# basic loops are 'for' loops
#counter in some vector

#looks like the if then, but not.
#going to go through the code block a bunch of times, once for each value in the vector. 
#fist time through, counter = 1, then 2, then etc

for(counter in 1:10) {
  print(counter)
}      



for(counter in c("here", "are", "some", "strings")) {
  print(counter)
}  

#where does 'counter' come from? we make it up. name of the variable that will get assigned a new value each time through the loop

values.to.print = c("a", "b", "c", "d")
for(counter in values.to.print) {
  print(counter)
}


values.to.print = c("a", "bee", "c", "dee")

for(counter in values.to.print) {    #behind the scenes: counter is "a", ... etc
  if(nchar(counter) == 1) {   #checking if length of the counter is equal to one
    print (counter)
  } else {
    break                 #gives up on the loop
  }
}

#note, though, that the value of counter is currently "bee". So you can print counter in your console to see where the loop broke if you're troubleshooting something. 



# going from copy-paste to a loop

#'what if we wanted to get a summary of sleep data.., of group1 and group 2'

mysleep = sleep
summary(mysleep)

summary(mysleep[mysleep$group ==1, ])
summary(filter(mysleep, group == 1))  #two ways to do this.

#now I want to see group 2. could copy paste and change to a 2

summary(mysleep[mysleep$group ==2, ])

#if you're copy/pasting and changing ONE thing - ideal for a loop

for(this.group in c("1", "2")) {
  print(summary(mysleep[mysleep$group ==this.group, ]))
  
}

#have to put explicit print statements in the loop!

for(i in c("1", "2")) {
  print(i)
}

#you name the variable! some people just call it i. scott thinks it's good to label it explicitly

```

# Some tips for loops

```{r}
# "Poor man's progress bar"
#might have loop that takes a long time to run....how can you know what's happening?

for(counter in 1:1e5) {
  if(counter %% 100 == 0) { cat(paste("Starting iteration", counter, "\n")) }
  mymean = mean(rnorm(1e6))
}

# sometimes when you're going through a loop, you want to set the counter equal to a string. below, might want to do a different histogram for each column in iris dataset.

# using get() to access objects programmatically
ggplot(iris, aes(Sepal.Width)) + geom_histogram()  #this works
ggplot(iris, aes("Sepal.Width")) + geom_histogram() #this does not work. it's a string
ggplot(iris, aes(get("Sepal.Width"))) + geom_histogram() #this works! get() gives you the object that the string is a name of.



# also you need explicit "print" statements inside loops

for(this.dim in colnames(iris)[1:4]) {
  
}

# also for plots!

for(this.dim in colnames(iris)[1:4]) {
  print(ggplot(iris, aes(get(this.dim))) + geom_histogram() +xlab(this.dim))

  }

```


# PRACTICE #3 & #4

# Building up to a power analysis via simulation
  - simulate fake data under some parameters
  - perform test
  - examine results of many iterations

```{r}
# quick review of d/p/q/r* family of distribution functions
?rnorm

x <- sort(rnorm(1000))
summary(x)
sd(x)
hist(x)
plot(x)
plot(dnorm(x) ~ x)
plot(pnorm(x) ~ x)
plot(qnorm(seq(0, 1, .01)) ~ seq(0, 1, .01))

2*(1-pnorm(abs(1.96)))

# simulating correlations with MASS
library(MASS) # look out for "masking" of select() from dplyr
cor.table <- matrix(c(1, .2, .2, 1), nrow = 2)
cor.table
some.data <- mvrnorm(n = 10, mu = c(1, 2), Sigma = cor.table, empirical = FALSE)
some.data
summary(some.data)
cor(some.data[, 1], some.data[, 2])
as.data.frame(some.data)


```

## Step 1: write the code for one basic "iteration" of a loop

```{r}
# simulate data


# perform test


```

## Step 2: figure out how you are going to store the results of each loop
  - "pre-allocate" an object to act as the "results container"
  - modify your one-time code to "fill in" part of that container

```{r}


```

## Step 3: put code into a loop structure and test it out with a few iterations
  - make a "progress bar" if you want

```{r}


```

## Step 4: run it "for real" with full # of iterations

```{r}


```


# Power analysis
  - set up parameters of simulation
  - run simulation loop
  - process/analyze results

```{r}


```

Going further: looping over sets of simulations!
  - loops within loops

```{r}
# simple loop within loop

# nested loops in power simulation

```

# Troubleshooting loops
  - build incrementally
  - "manually" set counter, step through code inside loop
  - if you hit a snag, use "last value" of counter to identify where you hit that snag

```{r}


```

# PRACTICE #4