---
title: "Day 8 Practice"
output: html_document
---

# Section 1

1. Write an if...else statement to test whether a string matches "right", and if so, print "you win!", but if the string matches "wrong", print "you lose!", and if it's neither, print "the result is inconclusive".

```{r}

x = "wrong"

if(x == 'right') {
  print('you win!')
} else if(x == 'wrong') {
  print('you lose!')
} else {
  print ('the result is inconclusive')
}


#scott's method

x = "right"

if(x %in% 'right') {
  print('you win!')
} else if(x %in% 'wrong') {
  print('you lose!')
} else {
  print ('the result is inconclusive')
}


#can do things with functions in here, use things like sapply to apply the complicated function to a whole bunch of values


# look at scott's final practice doc to get this.


```


2. Make a copy of the iris data. Use `ifelse()` to create a new column, where:
  - if the value of Sepal.Length is less than 5, the new column should have a value of "low"
  - if the value of Sepal.Length is greater than 6, it should have a value of "high"
  - and if it's between 5 and 6 (inclusive), it should have a value of "mid"
  - make a plot in ggplot2 to confirm that you did this right
  
```{r}

my.iris = iris

my.iris$new = ifelse(my.iris$Sepal.Length < 5,"low", "nope")

my.iris$new = ifelse(my.iris$Sepal.Length < 5,"low", 
                     ifelse(my.iris$Sepal.Length > 6, "high",
                            ifelse(my.iris$Sepal.Length <= 6 & my.iris$Sepal.Length >=5, "mid","")))


#don't really need the last ifelse with the >= <=, mid could also just be the 'no' argument.

my.iris$new = factor(my.iris$new, levels = c("low", "mid", "high"))

levels(my.iris$new)

ggplot(my.iris, aes(new, Sepal.Length)) + geom_boxplot()

ggplot(my.iris, aes(new, Sepal.Length)) + geom_boxplot() + ggtitle("Boxplot")


ggplot(my.iris, aes(Sepal.Length)) + geom_histogram()
ggplot(my.iris, aes(Sepal.Length)) + geom_histogram(aes(fill = new))

```


# Section 2

3. Practice writing a very simple loop, that simply prints out each level of Species from the iris data.  In other words, the result of running the loop should look something like:

"Level number 1 of Species is setosa"
"Level number 2 of Species is versicolor"
"Level number 3 of Species is virginica"

```{r}

#what do I want? need to loop SPECIES and SPECIES number

levels(my.iris$Species)

for(levelnum in 1:length(levels(my.iris$Species))) {
  print(levelnum)
}


for(levelnum in 1:length(levels(my.iris$Species))) {
  print(paste("Level number", levelnum))
}


for(levelnum in 1:length(levels(my.iris$Species))) {
  this.level = levels(my.iris$Species)[levelnum] #can create another variable within the loop
  print(paste("Level number", levelnum, "of Species is", this.level))
}


#can use cat() instead of print(), but then you have to add the line return

for(levelnum in 1:length(levels(my.iris$Species))) {
  this.level = levels(my.iris$Species)[levelnum] #can create another variable within the loop
  cat(paste("Level number", levelnum, "of Species is", this.level, "\n"))
}


# looping through strings (instead of through numbers, like above)

for(this.level in levels(my.iris$Species)) {
  levelnum = which(levels(my.iris$Species) %in% this.level)
  print(paste("Level number", levelnum, "of Species is", this.level))
}


```

4. Now using the loop from #3 as a starting point, for each pass through the loop, save a PDF of a scatterplot of Sepal.Length by Sepal.Width, but the plot should be different each time through the loop in the following ways:
  - each plot should only plot the subset of data from one species
  - each plot should display the plotted species as the plot title (either use `labs()` or `ggtitle()`)
  - each plot should be saved as a separate file, with the species in the file name
  - how could you change this code to write all three plots out to the same file?

```{r}
# ggtitle() example
ggplot(iris, aes(Sepal.Width)) + geom_histogram() + ggtitle("All species")


#do something once
pdf("all species plot.pdf")

for(levelnum in 1:length(levels(my.iris$Species))) {
  this.level = levels(my.iris$Species)[levelnum] #can create another variable within the loop
  this.data = my.iris[my.iris$Species %in% this.level, ]
  pdf(paste(this.level, "plot.pdf"))
  print(ggplot(this.data, aes(Sepal.Width)) + geom_histogram() + 
          ggtitle(this.level))
  dev.off()
}


#Can also - start pdf first, take out the pdf command in the loop, and put dev.off after the loop, to make one document with a new page for each plot

```

# Section 2

4. Similar to what we did with the correlation power simulations, write code (including a loop) that creates a power simulation for comparing the means of two normally-distributed samples.
  - compare the sample with both `t.test()` and `wilcox.test()`
  - put the p-values from both tests in each simulation result
  - compare the power of the t-test vs. the power of the Wilcoxon test
  
```{r}
# set up the parameters of the simulation
n.sim <- 1000
# what other parameters are needed?

# set up the "empty" data frame for the results
sim.results <- data.frame(# WHAT SHOULD YOUR COLUMNS BE? 
                          )

# start the loop
# each iteration of the loop should:
# 1. generate data
# 2. perform the analysis
# 3. store the results (here, p-values) in the results data frame


# examine the results
# create a column that indicates which of the p-values reached significance


# power = the proportion of "significant" results


```

