---
title: "Day 4 Homework"
output: html_document
---


1. Identify a column in your data that *should* be a factor (whether it is or not). Check to see whether R is already treating it as a factor, and if not, convert it into a factor.

```{r}
cvsiii = read.csv(file.choose())
attach(cvsiii)
summary(cvsiii)

#Contrast should be a factor - and is. 
#Subject should be a factor - and isn't.

cvsiii$Subject = as.factor(cvsiii$Subject)
summary(cvsiii$Subject)
```


2. What are the levels of the factor? How many occurrences of each level are there in the data? Are the levels ordered in a sensible/meaningful way?
   
```{r}

summary(AgeGroup)
levels(AgeGroup)
#this factor has 3 levels - YNH, ONH, OHI. I would like the factors to be ordered as such, but they are not. 

cvsiii$AgeGroup = factor(cvsiii$AgeGroup, levels = c("YNH", "ONH", "OHI"))
summary(cvsiii) 

```


3. Create a new column that is a numeric "version" of the factor column, using `as.numeric()`. Use `xtabs()` to see how the values of these columns line up.

```{r}
cvsiii$Age_numeric = as.numeric(AgeGroup)
summary(cvsiii)

xtabs(~Age_numeric, AgeGroup, data= cvsiii)
```


4. Create another column that is a copy of the factor column. Now, recode this new column so that the factor levels have new names. Use `xtabs()` or another function to double-check your coding.
  (e.g., if you have a factor with levels "A" and "B", you could make a new column that had "condition A" and "condition B" as the level names instead)

```{r}

cvsiii$Age_copy = cvsiii$AgeGroup
levels(AgeGroup)

levels(cvsiii$Age_copy) = c("Older HI", "Older NH", "Younger NH")
levels(cvsiii$Age_copy)
```


5. Change the order of the levels in your new factor, using three different methods:
  a. `relevel()`
  b. `reorder()`
  c. `factor()`

```{r}

cvsiii$Age_copy = relevel(cvsiii$Age_copy, "Older NH")
summary(cvsiii)

neworder = c("Younger NH", "Older HI", "Older NH")
cvsiii$Age_copy = reorder(cvsiii$Age_copy, neworder)     #not sure what to do about this error.

cvsiii$Age_copy = factor(cvsiii$Age_copy, levels = c("Younger NH", "Older NH", "Older HI"))
summary(cvsiii)

```


6. Missing values
  a. Do you have any missing values (NA) in your data?
  b. Are there any places where you *should* have missing values but they were read into R as some other value?
  c. If the answer to (b) is "yes", figure out how to read in your data in a way that correctly identifies the missing value "code"
  d. If you have any NAs in your data, describe why you think those values are missing. Does there appear to be any systematic relationship between missingness and other variables in your data?

```{r}

#no missing values

length(which (is.na(cvsiii) == TRUE))

cvsiii$nas = NaN

length(which (is.na(cvsiii) == TRUE))


```


7. Find some aspect of your data that needs to be combined or merged in some way.  This could be a variety of things:
  - multiple "sheets" in an Excel
  - multiple files that should be combined into one data set
  - properties that need to be merged in to one data set. For example:
    - you have a table of subject properties (gender, handedness, etc.) that need to be merged into the item-level data
    - you have other information on countries (population, etc.) that need to be merged in
  - if you DON'T have this naturally-occurring issue, figure out a way to manufacture it. For example:
    - split up your data into chunks
    - make up some new table of information that you want to merge in
  - Explain (in words) what you are trying to do
  - Use binding/merging commands to accomplish this
  - Complete at least ONE example of binding and ONE example of merging

```{r}

#I want to enter a column for age (in years) for each participant. But each participant has multiple lines of data, so Ill need to use some kind of conditional. (I don't think this will use bind or merge, and we haven't covered them yet, but I want to try it anyway.)

cvsiii$Years = NA
cvsiii$Subject = as.factor(cvsiii$Subject)
cvsiii$Years = ifelse(cvsiii$Subject == "1101", 28, NA)

#ok, this would take a while if you were doing it for each subject individually. #can I use cbind as a conditional of sorts? 

```
