---
title: "Day 4 Lecture"
output: html_document
---

# Today's agenda

  - Review & questions
  - Character/string data
  - Factors
  - Excursion: simple data frame manipulation, `xtabs()`
  - PRACTICE: working with factors, part 1
  - Special values: NA/NULL/Inf/NaN
  - PRACTICE: data type coersion
  - Lists
  - Creating data frames from scratch
  - Combining data frames
  - PRACTICE: binding & merging

# Review
  - Go over Day 3 Practice
  - Day 3 Homework example
  - Knitting notes

# Characters
  - more primitive than factors
  - can manipulate with regular expressions, other text functions

Some handy text stuff

```{r}
# quotes

x = "a string"
x
y = "also a string"
y

z = "this is a 'string' too"
z
w = 'this thing is also a "string"'
w

#can't do double quotes within double quotes

# cat(), print()

print(x) #returns a representation - reminding you that this is a vector
cat(x) #just prints the string out to the console - useful if you're doing something like printing a message out to the console and you just want it to look like text. 

cat(x, file = "cat_example.txt") #here, we printed to a file. it got saved to the working directory. 

sample.ttest = t.test(rnorm(20, 0, 1), rnorm(20,2,1))
sample.ttest
print(sample.ttest)
printed.ttest = print(sample.ttest)

# special characters: \t, \n, "escaping"  #basically these allow you to format the white space
x = "here's a cool result: 10"
cat(x)

x = "here's a cool result: \t10"
cat(x)

z = "here's a cool result: \n\n\t\t\t\t10"
cat(z)

cat("some\ntext")
cat("some\\ntext")  #double slash tells it to just treat the second slash like a character
cat("some\"text")

# nchar() #number of characters, length of a string
nchar("hello")


# paste() - putting strings into a bigger string.
paste("hello", "there")
x = c("here", "are", "some", "strings")
x
paste("prefix", x)

#collapse - wihtin paste, if you want to collapse vectors into a single string
paste (x, collapse = " ")

# strsplit()
y = paste(x, "suffix", sep = "_")

strsplit(y, split = "_")

z = paste(x, "suffix", sep = ".") #r will, for example, make white space in a column name a dot
z

strsplit(z, split = ".")  #but, dot is a special thing that means 'any character'. so this doesnt work 
strsplit(z, split = "\\." )


# stringr package

# grep() - searches for strings or patterns
x
grep("here", x)  #tells you where it found the string, the indices of the vector
grep("e", x)

# grepl() #logical version, returns trues and falses
          # could use this for logical subsetting

# gsub()  # search and replace function
          # have you give thing you're searching for, then replacement, then vector
gsub("e", "E", x)

#gsub("ONH", "oldernorm", cvsiii$AgeGroup) # this isn't working right now. not sure why. maybe because                                                  factors aren't strings? 
                  
                                            #update - correct. had to make the column as.character, then run                                             gsub. BUT also have to assign gsub to a new column

#cvsiii$AgeGroup = as.character(cvsiii$AgeGroup)
#cvsiii$Age_New = gsub("ONH", "oldernorm", cvsiii$AgeGroup)

#then will have to re-factor that new column in order to run whatever analyses. 

longer.thing = c("10F5657", "59J4957", "30K16238")
gsub(\\d+\\w(\\d+)", "\\1", longer.thing)

```


# Factors
  - Actually numbers under the hood
  - Act like numbers + labels
  - VERY useful for many kinds of analyses
  - if a column is a *property* of data (not a measure itself), factors are usually best
    - groupings = good as factors
    - conditions = good as factors
  - if a column is *data*, maybe consider keeping it a string
    - free response answers = bad as factors
  - `as.factor()` coerces to a factor
  - `factor()` creates a factor
  - factors have *levels*: categories
  - `stringsAsFactors` : option for how to read in character vectors
    - = FALSE if you don't want to read in characters as factors

```{r}
# factor()

x= c(rep("a", 20), rep("b", 20), rep("c", 20))
x
summary(x)

x.factor = factor(x)
summary(x.factor)

# levels()
levels(x.factor)

# summary()



# subsetting
mysleep = sleep
summary(mysleep)

class(mysleep$group)
levels(mysleep$group)


mysleep.group1 = mysleep[mysleep$group == '1',]
summary(mysleep.group1)

# droplevels    #gets rid of any factor levels that have 0 observations
mysleep.group1 = droplevels(mysleep.group1)
summary(mysleep.group1)

# changing levels (labels)
mysleep = sleep
levels(mysleep$group)

levels(mysleep$group) = c("pre", "post")
levels(mysleep$group)
summary(mysleep)

# re-ordering
mysleep = sleep
levels(mysleep$group) = c("pre", "post")
summary(mysleep)
levels(mysleep$group)

#how to make it so pre is the reference level?
mysleep$group2 = mysleep$group #created new column basically cloning the group
mysleepgroup2 = factor(mysleep$group2, levels = c("post", "pre"))



# coersion

```

# Data manipulation excursion
## Creating (simple) new columns

```{r}
# single value

mysleep$experiment = 1

mysleep$group3 = mysleep$group

# using subsetting

#ifelse()

mysleep$sleeplevel1 = ifelse(mysleep$extra >= 1, "high", "low") #if first condition is true, returns high, otherwise next thing

#mysleep$sleeplevel1 = ifelse(mysleep$extra >= 1, "high", ifelse(mysleep$extra <1, "low", NA))
#can next if elses


summary(mysleep)

mysleep$sleeplevel1 = as.factor(ifelse(mysleep$extra >= 1, "high", "low"))

summary(mysleep)


mysleep$sleeplevel1 = factor(ifelse(mysleep$extra >= 1, "high", "low"), levels = c("low", "high")) 
#creates the levels, as a factor, and specifies the order of the levels

summary(mysleep)


```

## Using xtabs() to check coding/cross-tabs

```{r}
#quickly double check a lot of things
xtabs( ~ group + group2, data = mysleep) #cross-tabulation of data. 
xtabs( ~ sleeplevel1 + extra, data = mysleep) #cross-tabulation of data. 

```

# PRACTICE: #0 -- #6

# NAs, NULLs

## NAs
  - missing data
  - can be a huge pain
  - hugely important to handle properly!
  - no universal "handle NAs" or "ignore NAs" argument
  - sometimes `na.rm = TRUE` (like for `mean()`)
  - sometimes it's other stuff
  - `cor()` use = "pairwise.complete.obs"
  - no universal "deal with NAs" argument
  
```{r}

```

## NULL
  - another special value
  - means "nothing" or "doesn't exist" (contrast with "missing" or "unknown")

```{r}

```

## Inf, -Inf, NaN
  - can sometimes result from mathematical operations
  - helpful to be able to identify, locate them

```{r}
1/0
```

# Data types, classes, & coersion

```{r}
# typeof()

# class()

```

`is.X` family of functions (`is.factor()`, `is.na()`, `is.finite()`, etc.)
`as.X` family of functions (`as.factor()`, `as.numeric()`, `as.data.frame()`, etc.)



# PRACTICE: #7

----------just do practice through number 7------------

# Lists
  - similar to vectors
  - can contain different data types
  - use [[ ]] to pull out values
  - handy for glomming large/complex objects together
  - data frames are special lists!
  
```{r}

```

# Creating data frames
  - handy, especially for making "look-up tables"
  
```{r}
# data.frame()

```

# PRACTICE: #8 -- #11

# Combining data frame objects
## `rbind()` and `cbind()`
  - rbind
    - glues rows on top of each other
    - expects same columns
    - new and improved `bind_rows()` in `dplyr` package
  - cbind
    - glues columns next to each other
    - need to be the same length    
```{r}

```

## `merge()` and `*_join()` from `dplyr` package
```{r}

```
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex
    - `nrow()` good check on what was gained/lost during a merge
    - "join" functions in `dplyr`
      - `inner_join()` is like all = FALSE
      - `full_join()` is like all = TRUE
      - `left_join()`, `right_join()`, etc.
      - see ?join for more details



# PRACTICE: #12 -- #14


