---
title: "Day 5 Lecture"
output: html_document
---

# Outline
  - Review Practice & Homework 4
  - Packages and namespace conflicts
  - Review NA & NULL
  - Lists
  - More on data frames
  - Simple transformations, "base" and "dplyr"
  - Simple data combinations: binding and merging
  - Graphics intro
  - REMINDER: project proposals due Sunday!

# Review
```{r}
#reminder of some 'handy functions'

#seq()
seq(from = 1, to = 10, by = 1)
seq(from = 20, to - 2, by = -2)

#rep()
rep(1, 10) #repeat 1 ten times

rep(1:3, 5)
rep(1:3, each = 5)
rep(1:3, 5, each = 2)
```



# Quick aside: package namespace conflicts

```{r}
library(MASS)
library(dplyr)

# select() from MASS is "masked"
# need to use MASS::select()

?MASS::select()
```

# More on NA, NULL

```{r}

x = c(1, 2, 56, NA, 43)
mean(x) #throws an NA, have to tell it to ignore the NA

mysleep = sleep
mysleep$extra[sample(1:nrow(mysleep), 5, replace = FALSE)] = NA #assigning NA to the values from these 5 random rows. simulating random missingness

summary(mysleep)

#what if we want to know where the NAs are?
mysleep_na = mysleep[is.na(mysleep$extra) , ]  ##doesn't work to do mysleep$extra == NA. because can't compare                                               a number to an unknown value! will throw NA for everything.

?complete.cases

mean(mysleep$extra, na.rm = TRUE)  #no generic function to specify remove na's. not always na.rm. differs by function

```


# Lists
  - similar to vectors
  - can contain different data types
  - use [[ ]] to pull out values
  - handy for glomming large/complex objects together
  - data frames are special lists!
  
```{r}
# list()

x = c(1, "two", FALSE) #c() makes it a vector, we get coersion, etc

y = list(1, "two", FALSE) 
y

      #for vectors, use single brackets to pull a single thing
x[1]

y[1] #each item in a list is still a list
y[[1]] #use double brackets to get the thing inside it - just the value


# unlist() #turns a list into a vector
x = "type_instance"
strsplit(x, split = "_")

unlist(strsplit(x, split = "_"))

# names() and str()
x = list(1, "two", FALSE)
names(x)
names(x) = c("first", "second", "third")

x[[2]]
x[["second"]]

# pulling apart t-test object

mytest = t.test(rnorm(20, 1), rnorm(20, 2))
mytest

names(mytest)
str(mytest)
mytest$p.value

#S3 vs S4 classes
#S3 - basic, S4 - fancy
#if you look at the bottom of the output structure, you see "class"0 = chr "htest"
#for S4 classes, you'll see an @ instead of  $


mysleep = sleep


nrow(mysleep)
length(mysleep)   #mysleep is just a fancy list. the numbers of the lists are columns. 

mysleep[["extra"]] #gives us back a vector
mysleep[, "extra"]
mysleep["extra"] #here, still a list (still a dataframe)

```

# Creating data frames
  - handy, especially for making "look-up tables"
  
```{r}
# data.frame()

my.df = data.frame(var1 = rnorm(20,0,1), 
           var2 = rnorm(20,0,2), 
           missing = NA, 
           condition = rep(c("A", "B"), each = 2))

summary (my.df)

# stringsAsFactors argument

my.df.nofactors = data.frame(var1 = rnorm(20,0,1), 
                             var2 = rnorm(20,0,2), 
                             missing = NA, 
                             condition = rep(c("A", "B"), each = 2), stringsAsFactors = FALSE)

summary(my.df.nofactors)
```

# Intro to `dplyr`

```{r}
#dplyr is all about manipulating data
library(dplyr)

```

## subsetting with `dplyr`

```{r}

mysleep = sleep
mysleep.group1 = mysleep[mysleep$group %in% "1", ] #old way
class(mysleep.group1)

mysleep.group1.dplyr = filter(mysleep, group %in% "1") #dplyr way. filter() takes a conditional

mysleep.group1.dplyr = mysleep %>% filter(group %in% "1") #other dplyr way? something when you put things through dplyr, you get structures fancier than plain data frames, like tibbles. 



mysleep.group2 = mysleep[mysleep$ID %in% c("1", "4", "7"), ]


```

## "tibbles" and `as.data.frame()`

```{r}

```

# Simple transformations    

## Creating new columns

```{r}

mysleep = sleep
mysleep$extra2 = mysleep$extra #creates a new column with the values currently existing in 'extra', base R

#mutate() from dplyr
mysleep = mutate(mysleep, extra3 = extra) #dplyr way uses mutate()
head(mysleep)


```

## Simple column transformations

```{r}
mysleep = sleep
mysleep$doubleextra = mysleep$extra * 2
head(mysleep)

# easy way to do things like log transforming

mysleep$logextra = log(abs(mysleep$extra))
head(mysleep)

#dplyr style - also uses mutate

mysleep = mutate(mysleep, logextra2 = log(abs(extra)))
head(mysleep)

```


# Combining data frame objects
## `rbind()` and `cbind()`
  - rbind
    - glues rows on top of each other
    - expects same columns
    - new and improved `bind_rows()` in `dplyr` package
  - cbind
    - glues columns next to each other
    - need to be the same length    
```{r}
mysleep = sleep #"here's a case where factors are going to be annoying"
summary(mysleep)
#let's say we've collected data on two more people
newsleep = data.frame(extra = rnorm(4), 
                      group = rep(c("1", "2"), each = 2),
                      ID = rep(c("11", "12"), 2))
newsleep

allsleep1 = rbind(mysleep, newsleep)
summary(allsleep1)

newsleep2 = data.frame(extra = rnorm(4), 
                      group = rep(c("1", "2"), each = 2),
                      ID = rep(c("11", "12"), 2), 
                      gender = rep(c("M", "F"), 2))


allsleep2 = rbind(newsleep, newsleep2)
summary(allsleep2)


allsleep3 = bind_rows(mysleep, newsleep2) #throws errors because mismatching character/factor. be careful!
summary(allsleep3)
```


## `merge()` from `base` R, and `*_join()` from `dplyr` package
```{r}
mysleep = sleep

#lets say we have table of age info, 

sleepages = data.frame(ID = as.character(1:10), 
                       age = sample(20:50, 10))
sleepages

#now we want that age data in the sleep data frame to do correlations or whatever. 

#must have some sort of linking variable - allows matching of two dataframes.

sleepmerged = merge(mysleep, sleepages) #this is a simple situation! same column names, same values, so on.
sleepmerged


sleepages2 = data.frame(ID = as.character(3:12), #what happens if we make it mismatch a bit??
                       age = sample(20:50, 10))
sleepmerged2 = merge(mysleep, sleepages2) #this is a simple situation! same column names, same values, so on.
sleepmerged2    #now we get a subset - no data for 11 and 12, and people 1 and 2 disappear. need the argument called 'all' *it's false by default* - easy to forget. 

sleepmerged3 = merge(mysleep, sleepages2, all = TRUE)
sleepmerged3        #keeps everything and sticks in NAs


#dplyr way uses *_join
sleep.inner = inner_join(mysleep, sleepages2, by = "ID") #same as first - drops the non-overlap
sleep.inner

sleep.full = full_join(mysleep, sleepages2, by = "ID") #gives everything
sleep.full

sleep.left = left_join(mysleep, sleepages2, by = "ID") #starts with the *left* dataframe (mysleep), keeps that and appends as much as possible from the other.
sleep.left



```
    - simplest: have ONE column in common, named the same thing (with the same values)
    - can get MUCH more complex
    - `nrow()` good check on what was gained/lost during a merge
    - "join" functions in `dplyr`
      - `inner_join()` is like all = FALSE
      - `full_join()` is like all = TRUE
      - `left_join()`, `right_join()`, etc.
      - see ?join for more details

# PRACTICE

# Intro to graphics

http://accidental-art.tumblr.com/

## The importance of graphing

```{r}
?anscombe
example("anscombe")


```

## Two systems

  - "base" graphics
  - "grid" graphics
  - See *R Graphics* by Paul Murrell for a very thorough "deep-dive" into these implementations:  https://www.stat.auckland.ac.nz/~paul/RG2e/

### Base graphics
  - more "primitive"
  - can also be much faster than grid
  - lots of isolated plot functions
  - lots of weird little parameters to worry about
  - some handy quick-plot functions from base graphics:
    - `plot()` : a generic function that will often produce something interesting, can do scatterplots with a formula `y ~ x`
    - `hist()` : histogram of a vector
    - `boxplot()` : boxplot, using a formula `variable ~ grouping.factor`

### Grid graphics
  - better for more "high-level" plots
  - much more complex underlying system (IMHO)
  - more powerful and extendable
  - the basis of extremely popular packages such as `lattice` and `ggplot2`

## Graphics in other packages
  - many packages will have some kind of graphics functions in them, for visualizing specific kinds of things
  - these will often be in base graphics, "pre-canned", etc.
  - the two best more general-purpose graphics packages are `lattice` and `ggplot2`
  - `ggplot2` is another Hadley Wickham project, and has the most documentation, examples, and is currently most widely used

# Using `ggplot2`

## A "Grammar of Graphics"
  - based on work by Leland Wilkinson, a "theory" of graphic display of data
  - Hadley implemented this framework as his dissertation
  - `ggplot2` is the overhauled version of the original implementation, and it is still under active development
  - the idea is to implement a general system for mapping data variables to *aesthetics* (Wilkinson's term), i.e., visual dimensions (x, y, color, etc.)
  - allows for high-level control of the structure of a plot, to allow for flexible, powerful visualizations
  - the place for documentation: https://ggplot2.tidyverse.org/reference/
  - another good site: http://www.cookbook-r.com/Graphs/
  
## Basic structure and concepts

  - unlike base graphics, the *value* of `ggplot2` objects is a complex object, and `print()`-ing them renders the graph.
  - the value of the object is built up by adding *layers*
  - start with a call to `ggplot()` that specifies the data and some mappings of variables to aesthetics
  - add "geoms" to actually render something
  - add other things for fine control over graphical parameters, etc.

## Exploring data with basic geoms and aesthetics

```{r}
library(ggplot2)
?starwars

sw = starwars
head(sw)



# geom_histogram()
ggplot(sw, aes(mass)) + geom_histogram()


# geom_density()

# geom_point()

ggplot(sw, aes(mass, height)) + geom_point()

# geom_text()
ggplot(sw, aes(mass, height)) + geom_text(aes(label = name))


# geom_boxplot()

```

## More on colors
  - http://colorbrewer2.org/

```{r}
colors()
?RGB

# scale_color_brewer()

```

### Saving plots

```{r}
?Devices

```

TIP: look into other PDF readers (not Adobe) if you want to leave the file "open" (i.e., displayed) on your computer while you are editing/changing the file


